How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985

### 2022-05-20 Draft_May_2022/Full_draft_20220510.rmd
Added latest draft version to rmd file

---
title: How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article; names are ordered alphabetically. Both of them are PhD candidates at The Geneva Graduate Institute. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Instituteis affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute affiliated to the Centre for International Environmental Studies.]
date: "May 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document:
    reference_docx: template.docx
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon Rainforest is one of the most important ecosystems in the world, and most of this biome is located in Brazil. Albeit the importance of understanding the relationship between the social construction of the Amazon and its effects on environmental outcomes, we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Discursive problem-construction within democracies legitimize ways of thinking and acting towards the environment while affecting decisions to deforest, or not, at the local level. Building on Hirschman’s conceptualization of chosen problems (1963), we propose a framework to investigate how the Brazilian Amazon has been constructed as a problem in political discourse. Combining hand-coding and supervised machine learning, we classify statements about the Amazon forest or region in 6130 official presidential speeches since 1985. We find that, first, international environmental events appear to drive presidents to speak more about the Amazon. Second, while the Amazon as a problem of economic integration dominated discourses from 1985 to the mid-2000s, environmental conservation and social development constructions briefly surpasses economic integration from 2010 to 2015. Constructing the Amazon as an issue of sovereignty also became more pertinent from 2010 onwards. Finally, using a multinomial regression model, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation, than economic integration or social development, as presidents move away from the Amazon region.
---

@@

                                      mixed_type == '0_1_1_0' ~ "Economic and Social Development",

@@

                                      mixed_type == '0_1_1_1' ~ "Sustainable Development",
                                      mixed_type == '1_1_0_0' ~ "Sovereign Integration",

@@

                    "Pure Sovereignty", "Other", "Economic Conservation", "Economic and Social Development",
                    "Sovereign Conservation","Sustainable Development","Sovereign Integration",
                    "Social Conservation", "Other Mixed-types")),

@@

                             grepl("Economic Conservation|Economic and Social Development|
                                   |Sovereign Conservation|Sustainable Development|
                                   |Sovereign Integration|Social Conservation|Other Mixed-types",

@@

Different governments along Brazilian history have been described as proponents of a specific view of the Amazon (Drummond & Barros-Platiau 2016; Padua 2012; Franchini & Viola 2019; Capobianco 2019; Pereira & Viola 2021). For example, while the military dictatorship is associated with both national sovereignty and then economic integration, President Sarney and President Lula are often tied to environmental conservation. However, governments are usually represented monolithically, advancing a view that specific governments see the Amazon as an instance of one specific problem. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes (Waroux et al. 2021), we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between and within governments.

In this article, we investigate how the Brazilian Amazon has been constructed as a problem in political discourses. Building on Hirschman’s concept of chosen problems (1963) and Bacchi’s concept of problem-representation in policy (2009), we propose a framework to identify problem-constructions in political discourses. Although problem-construction takes place in a series of instances, we analyze the case of discourses by Brazilian presidents since 1985. We opt for presidential speeches for two reasons. First, presidential discourses have the power to introduce and justify public policy, as well as shape its perception to broad audiences (Zarefsky 2004). It legitimizes ways of thinking about the Amazon. In turn, policy perception is key for policy adoption and implementation (Alesina and Giuliano 2009; López et al. 2020). This is especially pertinent as deforestation rates in Brazil are associated with expectations from government, generated from material and discursive governmental practices (Assunção, Gandour, and Rocha 2015; Capobianco 2019, 2021), and are a crucial factor in decisions to deforest at the ground (Campbell 2015). Second, we argue that presidents can employ specific problem-constructions that build objects as specific problems depending on where they are. Presidential discourses take place in different sites, from launching a new bridge in a small municipality in the middle of the Amazon to a keynote speech in a business association in São Paulo or at the UN general assembly in New York. Working with presidential discourses allows us to identify this variation in meaningful ways and better how the Amazon is socially constructed.

To investigate how the Brazilian Amazon has been constructed as a problem in political discourses, we create a dataset containing 6130 official presidential speeches by all Brazilian presidents since 1985. We subset the dataset by identifying Amazonian related statements within these speeches. We find that 2014 sections in these discourses refer to the Amazon at least once. We then develop a codebook grounded on Amazonian historiography to code how each of these statements constructs the Amazon as a particular problem. We use this codebook to manually code a randomly selected training set of the Amazonian related statements. We train a supervised machine-learning model in the hand-coded set and automatically label the remaining set of Amazonian statements. We then conduct a descriptive and inferential analysis of this data, tying our findings to endogenous and exogenous events related to deforestation.

We find that, first, international environmental events appear to drive presidents to speak more about the Amazon. Second, while the Amazon as a problem of economic integration dominated discourses from 1985 to the mid-2000s, environmental conservation and social development constructions briefly surpasses economic integration from 2010 to 2015. Constructing the Amazon as an issue of sovereignty also became more pertinent from 2010 onwards. Finally, using a multinomial regression model, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation, than economic integration or social development, as presidents move away from the Amazon region.

@@

Governments often have a choice in how they construct problems and what policy solutions they implement. In “Journey towards Progress”, Hirschman (1963) draws a conceptual distinction between pressing problems (pressured from outside parties to the government) and chosen problems (chosen by the government at their own discretion). Pressing problems can be either privileged or neglected, depending on the degree of pressure exercised by the interested groups. Chosen problems are those governments select at their discretion. Problems can change from pressing to chosen across time and in space as a function of (a) solutions becoming available, (b) a change of level of government control in society, or (c) a shift of interests from top policymakers (Hirschman 1963, 388–91). As Bacchi argues, though, policies have a cultural dimension as “it takes shape within specific historical and national or international contexts”. (2009, 10). The existence or proposal of a policy implies that there is a (public) problem that needs (governmental) action to be fixed. The alleged problem is not always explicitly stated in policy. Hirschman exemplifies chosen problems with the case of the construction of Brasilia (1975, 388). But building Brasilia can solve a problem of regional inequality, a problem of a dormant economy without state investment, a problem of political representation, or all three. It is up to the discretion of the government to highlight or not the implicit problem a policy solves.

Governments can choose to emphasize (or not) one or more implicit problems that a policy solves depending on which interest group they are in communication with. Relative to how the policy is represented to be, it can be a solution of problems that are considered pressing or not for different groups and it is up to the discretion of the political actor to construct a particular problem in a particular way given context. Putnam’s (1988) seminal article on the two-level game can help us make sense of this variation. The author argues that the outcomes of international negotiations lie within the overlap between the pressure of domestic groups and international groups. Taking the example of the 2012 Forest Code and the Amazon, President Dilma could emphasize its rural cadaster for environmentalists at the Paris Summit, or the amnesty for deforestation for agricultural elites.  The same policy, then, can solve the problem of environmental conservation as well as economic integration.

This variation in problem-construction suggests that governments are more diverse in their positions as the literature suggests. Several scholars describe federal governments' as proponents of specific policies for the Amazon Forest and region, using the vocabulary of policy cycles, periods, or phases (see Drummond & Barros-Platiau 2016; Padua 2012; Franchini & Viola 2019; Capobianco 2019; Pereira & Viola 2021). In this literature, the 1964 military dictatorship is associated with securing national sovereignty in the region by populating it and integrating it to the national economy (Drummond & Platiau 2016). The governments from the late 1980s up to 2009, are more associated with environmental conservation. The presidencies of Dilma and Temer are connected to the decline of environmental policies, while Bolsonaro to the complete dismantling of environmental policies. While these works are important to understand how different governments acted towards the Amazon, the classification of specific governments into periods or policy cycles represents them as monolithic: they associate specific governments to one specific view of the Amazon. The presidency of Lula, often associated with environmental conservation because we saw an unprecedented decrease in deforestation. At the same time, during his terms governmental credits to agricultural expansion and the cattle industry surpassed BRL 578 billion (Capobianco, 2021, 106). This sum is about 20 times more than the budget of the Ministry of the Environment for this entire period. There is more variation within governments than the literature suggests.

An interesting site to study the relationship between varied problem-construction and policy issues is presidential speeches as they legitimize ways of thinking about an issue and. While the relationship between presidential speeches and policy is not causal (Lawrence 2004), they have the power to introduce and justify public policy, as well as shape its perception to broad audiences (Zarefsky 2004). Even if presidential discourse might not persuade public opinion regarding a proposed solution for a problem (Eshbaugh-Soha 2010; Rottinghaus 2009; Krebs and Jackson 2007), they direct the range of what is included in national debates (Druckman and Holmes 2004). As presidents speak in varied locations to varied audiences, it is very likely that they chose to frame similar issues differently.

The possibility of different problem-constructions at the level of presidential speeches has important consequences for democracy. If presidents can promote different problem-constructions in different places, it is important to understand to what extent the implemented agenda correlates with their local, federal, or international problem-constructions. Pacheco (2019) proposes that we see the Amazon frontier as a key analytic category to understand the Brazilian state and democracy. Specifically, the author states that the natural richness of the region has been instrumentally transformed in political support through resource exploration by different governments over the last centuries. The costs for said economic benefits are the livelihoods of indigenous and traditional populations and the ecosystems they reside in. Political stability, thus, can be seen as a product of the trade-off between both. Policies during the military dictatorship were strongly geared toward integrating the Amazon into the national territory and international economy. These were times when deforestation increased, but political stability was achieved as those bearing the costs of said deforestation were being targeted. Generally high levels of violence in the region, epitomized in the assassination of Chico Mendes, serve as evidence for that (Hecht and Cockburn 1990). With the return of democracy in the 1980s, violence could not be used at the same extent as before and policies had to respond to growing environmental concern.

Relatedly, the possibility of different problem-constructions has consequences for environmental outcomes as well. When Brazilian presidents speak about the Amazon it not only makes headlines, nationally and internationally (Brice and Smith 2021; Harris 2021; Miranda 2021), but also incites responses, shapes expectations, and feeds into the behavior of many actors involved in the Amazon. Deforestation rates in Brazil are more responsive to the government’s environmental policy than exogenous factors as market fluctuations (Assunção, Gandour, and Rocha 2015; Capobianco 2019, 2021). Ethnographic research has also shown that policy expectations, generated from material and discursive governmental practices, are a crucial factor in decisions to deforest at the ground (Campbell 2015).

@@

Closely related to economic integration, a problem-construction that is often neglected while describing policy cycles is social development. Governments can emphasize the lack of hospitals, sanitation, and schools concerning peoples’ dignity, standards of living, and constitutional rights when speaking about the Amazon. This is not only relevant as a construction, but speaks, minimally, to democratic concerns over providing electorates, especially in the Amazonian region, with better standards of living than previously.

@@

To analyze how presidents construct the Amazon as a problem in time, we build upon the dataset provided by (Cezar 2020) which contains all official speeches by Brazilian Presidents from 1985 to 2019 scrapped from the archives of the Brazilian Presidential Library. We update the dataset by scraping and adding all official speeches from 2020 and 2021. The final dataset encompasses 6130 speeches for all the presidents of Brazil since 1985. Then, we proceed to identify all speeches that refer to the Amazon as a region, people, or forest out of the 6130 speeches. We do so by detecting all speeches in which the stem “amazon” appears. In Portuguese, the stem captures terms such as “Amazonia”, “Amazonica”, “Amazonidas”, “Amazonense(s)”, “Amazonas”, among others. We find that 946 speeches are, at least partially, about the Amazon from the 6130.

@@

We develop a codebook to code Amazonian statements in one or more problem-constructions (see codebook in appendix). In their conceptualization, each problem-construction is mutually exclusive, meaning that they cover different forms of constructing the Amazon as a problem. Nevertheless, each Amazonian statement might be assigned to one or more codes. A statement can, for example, construct the Amazon as a problem of sovereignty and a problem of economic integration, or a problem of social development and conservation. Amazonian statements, thus, can be either coded as pure-types or mixed-types. Mixed-types are relevant as constructions and policies are often multifaceted while portraying the Amazon as a combination of these problem-constructions. Figure 1 portrays this operationalization strategy.

```{r figure1, echo=FALSE, fig.cap="Figure 1: Operationalization of Problem-constructions", out.width = '100%'}
knitr::include_graphics("figure1.png")
```

@@

We chose to employ a support-vector machine (SVM) algorithm to label texts, that is, a non- probabilistic linear classifier that classifies documents by assigning points in mapped space to maximize the gap between binary categories [Meyer et al. (2021); Noble (2006)] . The SVM model is first trained using the hand coded training set and then employed to classify observations in the hand coded validation set. The trained SVM model was, on average 82%, accurate in classifying observations in the validation set before being tuned. After the SVM model is tuned, we use the model to automatically code the remaining 1007 Amazonian statements. The final dataset for analysis, excluding false positive matches, contains 1895 coded Amazonian statements. Finally, we extract locations for all speeches in the data. These locations represent the Brazilian state in which certain speech was given or an international country.

@@

Mentions of the Amazon across official presidential speeches vary in time. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches each year. We observe various local maxima: 1989, 1992, 2009, 2015, and 2019. These points, usually, coincide with events that help us explain the rises and falls of the Amazon in presidential discourse in time. For example, in 1985 the Amazon appeared in about 4% of all the presidential speeches. By 1989, the Amazon appeared in 32% of all speeches. This is the period when the Brazilian Constitution was written. Indigenous and traditional populations were instrumental in advocating for constitutional environmental rights and the protection of their territories (Hecht and Cockburn 1990). These were eventually enshrined in article 225, which gives all Brazilians a right to a balanced environment, and in article 231, which grants indigenous and traditional populations a right over their territory. Besides,  other factors are likely to explain this increase: in 1988 Chico Mendes was brutally murdered and the New York Times published an article with pictures of the Amazon burning (Simons and Times 1988). Both incidents caught unprecedented international attention and Sarney responded to these publicly with a set of policies to address deforestation (Capobianco 2021).

```{r Figure 2: Amazonian speeches by year, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, include=TRUE}

@@

  geom_text(aes(x = 1988, y = 0.34, label = "Chico Mendes/NYT"), size = 3) +

@@

  geom_vline(xintercept = c(1988, 1992, 2009, 2015),

@@

In 1990 the Amazon appeared in 12% of all speeches and, by 1992, this increased to 25% of all speeches. The 1992 Earth Summit, held in Brazil, and prepared by various state and non-state actors in the region, brought international attention to environmental topics. One of the big announcements, for example, was the consolidation of the first transnational partnership for the Amazon, the G7 Pilot Programme, which brought a high number of financial resources to the region for public policy implementation (Capobianco 2021). Throughout the 1990s and 2000s, however, the Amazon appeared, on average, on 15% of all speeches. By the end of the 2000s, we see another spike in the Amazon appearing in official speeches. This spike coincides with the 2009 Copenhagen Summit as well as the steepest decrease in deforestation rates since data is available. Lula led the delegation to Copenhagen with a self-image of “we do not promise, we deliver” when the stakes about climate change were high (Franchini and Viola 2019).

From 2010 to 2018, with the exception of 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. The Amazon appeared in 8% of official speeches in 2014. In 2015, the Amazon appeared in 19% of speeches.  This spike coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Disagreements related to the priority of environmental preservation over economic development, most notably in the case of Itaipu Dam, led the environmental minister Marina Silva to resign and run for the presidency on her agenda in 2010. These are also the years when Brazil entered a long period of political and economic instability, which led to the impeachment of Dilma Roussef. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen, and a perception that there was a turn towards less conservation since Marina Silva left the government and the 2011 New Forest Code was approved. By 2016, however, mentions of the Amazon in official speeches went down to 6%, the lowest share since 1985.

We subsequently observe a steady increase from 6% in 2016 to almost 24% in the first year of Bolsonaro’s presidency, 2019. At the time, international and national media brought unprecedented attention to Brazilian environmental issues as the record burning of the Amazon, and of the red sky afternoon in São Paulo circulated on social media and international media outlets (cite). This brought environmental issues as high priorities in Bolsonaro’s governing agenda. At the same time, President Bolsonaro retrieves Brazil’s hosting status for COP25 and begins to dismantle Brazilian environmental governance (cite).

@@

We conceptualize four pure problem constructions: sovereignty, economic integration, social development, and conservation. At the level of the speech, though, presidents might mix multiple problem construction in an Amazonian statement. Pure problem constructions dominate, average around 56% of all amazonian statements across time. As above, we observe a strong variation over time for most of these constructions. Figure 3, below, illustrates the share of each pure type in time.

>>>>

---
title: How has the Brazilian Amazon been constructed as a problem? The economy, the environment, the people, and the nation in the presidential speeches since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article; names are ordered alphabetically. Livio is a PhD Candidate in Anthropology and Sociology at the Geneva Graduate Institute and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences at the Geneva Graduate Institute and affiliated to the Centre for International Environmental Studies.]
date: "20 May 2022"
output:
  pdf_document: default
  fig_caption: yes
  html_document:
    df_print: paged
  word_document:
    reference_docx: template.docx
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon Rainforest is one of the most important ecosystems in the world, and most of this biome is located in Brazil. Albeit the importance of understanding the relationship between the social construction of the Amazon and its effects on environmental outcomes, we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments. Discursive problem-construction within democracies legitimize ways of thinking and acting towards the environment while affecting decisions to deforest, or not, at the local level. Building on Hirschman’s conceptualization of chosen problems (1963), we propose a framework to investigate how the Brazilian Amazon has been constructed as a problem in political discourse. Combining hand-coding and supervised machine learning, we classify statements about the Amazon forest or region in 6130 official presidential speeches since 1985. We find that, first, international environmental events appear to drive presidents to speak more about the Amazon. Second, while the Amazon as a problem of economic integration dominated discourses from 1985 to the mid-2000s, environmental conservation and social development constructions briefly surpasses economic integration from 2010 to 2015. Constructing the Amazon as an issue of sovereignty also became more pertinent from 2010 onwards. Finally, using a multinomial regression model, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation, than economic integration or social development, as presidents move away from the Amazon region.
---

@@

                                      mixed_type == '0_1_1_0' ~ "Social Economic",

@@

                                      mixed_type == '0_1_1_1' ~ "Economic, Social,\nand Conservation",
                                      mixed_type == '1_1_0_0' ~ "Economic Sovereignty",

@@

                    "Pure Sovereignty", "Other", "Economic Conservation", 
                    "Social Economic", "Sovereign Conservation", "Economic, Social,\nand Conservation",
                    "Economic Sovereignty", "Social Conservation", "Other Mixed-types")),

@@

                             grepl("Economic Conservation|Social Economic|Sovereign Conservation|
                                   |Economic, Social,\nand Conservation|Economic Sovereignty|
                                   |Social Conservation|Other Mixed-types",

@@

Different governments along Brazilian history have been described as proponents of a specific view of the Amazon (@Drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021). While the military dictatorship is associated with both national sovereignty and then economic integration, President Sarney's presidency is often tied to environmental conservation. However, governments are usually represented monolithically, advancing a view that specific governments see the Amazon as an instance of one specific problem. Albeit the current calls to understand the environment as a social-cultural construction and to identify the effect of culture on environmental outcomes (@lepolaindewaroux2021), we lack empirical accounts of how the Brazilian Amazon has been constructed as a problem in discourses over time, by geographical location, and between, or within, governments.

In this article, we investigate how the Brazilian Amazon has been constructed as a problem in political discourses. Building on Hirschman’s concept of chosen problems (1963) and Bacchi’s concept of problem-representation in policy (2009), we propose a framework to identify problem-constructions in political discourses. Although problem-construction takes place in a series of instances, we analyze the case of discourses by Brazilian presidents since 1985. We opt for presidential speeches for two reasons. First, presidential discourses have the power to introduce and justify public policy, as well as shape its perception to broad audiences (@zarefsky2004). It legitimizes ways of thinking about the Amazon. In turn, policy perception is key for policy adoption and implementation (@alesina2009; @lopez2020). This is especially pertinent to deforestation in Brazil as expectations from governments’ response, generated from material and discursive governmental practices (@assuncao2015; @capobianco2019; @capobianco2021), are a crucial factor in decisions to deforest at the ground (@campbell2015). Second, we argue that presidents can employ specific problem-constructions that build objects as specific problems depending on who they speak to. Presidential discourses take place in different sites, from launching a new bridge in a small municipality in the middle of the Amazon to a keynote speech in a business association in São Paulo or at the UN general assembly in New York. Working with presidential discourses allows us to identify this variation in meaningful ways and better understand how the Amazon is socially constructed.

To investigate how the Brazilian Amazon has been constructed as a problem in presidential discourses, we create a dataset containing 6130 official presidential speeches by all Brazilian presidents since 1985. We subset the dataset by identifying Amazonian related statements within these speeches. We find that 2014 sections in these discourses refer to the Amazon at least once. We then develop a codebook grounded on Amazonian historiography to code how each of these statements constructs the Amazon as a particular problem. We use this codebook to manually code a randomly selected training set of the Amazonian related statements. We train a supervised machine-learning model in the hand-coded set and automatically label the remaining set of Amazonian statements. We then conduct a descriptive and inferential analysis of this data.

Our findings are threefold. First, we find that international environmental events appear to drive presidents to speak more about the Amazon. Second, while constructing the Amazon as a problem of economic integration dominated presidential discourses from 1985 to the mid-2000s, environmental conservation and social development constructions steadily grew from 2000s onwards briefly surpassing economic integration as more common problem constructions from 2010 to 2015. Unprecedentedly, constructing the Amazon as an issue of sovereignty became more pertinent from 2010 onwards. Finally, using a multinomial regression model, we find that presidents are more likely to construct the Amazon as a problem of environmental conservation, than economic integration or social development, as presidents move away from the Amazon region.

@@

Governments often have a choice in how they construct problems and what policy solutions they implement. In “Journey towards Progress”, @hirschman1963 draws a conceptual distinction between pressing problems (pressured from outside parties to the government) and chosen problems (chosen by the government at their own discretion). Pressing problems can be either privileged or neglected, depending on the degree of pressure exercised by the interested groups. Chosen problems are those governments select at their discretion. Problems can change from pressing to chosen across time and in space as a function of (a) solutions becoming available, (b) a change of level of government control in society, or (c) a shift of interests from top policymakers (@hirschman1963, p. 388–91). As @bacchi2009 argues, though, policies have a cultural dimension as “it takes shape within specific historical and national or international contexts”. (p. 10). The existence or proposal of a policy implies that there is a (public) problem that needs (governmental) action to be fixed. The alleged problem is not always explicitly stated in policy. @hirschman1975 exemplifies chosen problems with the case of the construction of Brasilia (p. 388). But building Brasilia can solve a problem of regional inequality, a problem of a dormant economy without state investment, a problem of political representation, or all three. It is up to the discretion of the government to highlight or not the implicit problem a policy solves.

Governments can choose to emphasize (or not) one or more implicit problems that a policy solves depending on which interest group they are in communication with. Relative to how the policy is represented to be, it can be a solution of problems that are considered pressing or not for different groups and it is up to the discretion of the political actor to construct a particular problem in a particular way given context. Putnam’s (1988) seminal article on the two-level game can help us make sense of this variation. The author argues that the outcomes of international negotiations lie within the overlap between the pressure of domestic groups and international groups. Taking the example of the 2012 Forest Code and the Amazon, President Dilma could emphasize its rural character for environmentalists at the Paris Summit, or the amnesty for deforestation for agricultural elites.  The same policy, then, can solve the problem of environmental conservation as well as economic integration.

This variation in problem-construction suggests that governments are more diverse in their positions as the literature suggests. Several scholars describe federal governments' as proponents of specific policies for the Amazon Forest and region, using the vocabulary of policy cycles, periods, or phases (see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021). In this literature, the 1964 military dictatorship is associated with securing national sovereignty in the region by populating it and integrating it to the national economy (@drummond2006). The governments from the late 1980s up to 2009, are more associated with environmental conservation. The presidencies of Dilma and Temer are connected to the decline of environmental policies, while Bolsonaro to the complete dismantling of environmental policies. While these works are important to understand how different governments acted towards the Amazon, the classification of specific governments into periods or policy cycles represents them as monolithic: they associate specific governments to one specific view of the Amazon. The presidency of Lula, often associated with environmental conservation because we saw an unprecedented decrease in deforestation. At the same time, during his terms governmental credits to agricultural expansion and the cattle industry were 20 times more than the budget of the Ministry of the Environment for this entire period (@capobianco2021, p. 106). That is, there is more variation within governments than the literature suggests.

An interesting site to study the relationship between varied problem-construction and policy issues is presidential speeches as they legitimize ways of thinking about an issue and. While the relationship between presidential speeches and policy is not causal, they have the power to introduce and justify public policy, as well as shape its perception to broad audiences (@zarefsky2004). Even if presidential discourse might not persuade public opinion regarding a proposed solution for a problem (Eshbaugh-Soha 2010; Rottinghaus 2009; Krebs and Jackson 2007), they direct the range of what is included in national debates (Druckman and Holmes 2004). As presidents speak in varied locations to varied audiences, it is very likely that they chose to frame similar issues differently.

The possibility of different problem-constructions at the level of presidential speeches has important consequences for democracy. If presidents can promote different problem-constructions in different places, it is important to understand to what extent the implemented agenda correlates with their local, federal, or international problem-constructions. @pacheco2019 proposes that we see the Amazon frontier as a key analytic category to understand the Brazilian state and democracy. Specifically, the author states that the natural richness of the region has been instrumentally transformed in political support through resource exploration by different governments over the last centuries. The costs for said economic benefits are the livelihoods of indigenous and traditional populations and the ecosystems they reside in. Political stability, thus, can be seen as a product of the trade-off between both. Policies during the military dictatorship were strongly geared toward integrating the Amazon into the national territory and international economy. These were times when deforestation increased, but political stability was achieved as those bearing the costs of said deforestation were being targeted. Generally high levels of violence in the region, epitomized in the assassination of Chico Mendes, serve as evidence for that (@hecht1990). With the return of democracy in the 1980s, violence could not be used at the same extent as before and policies had to respond to growing environmental concern.

Relatedly, the possibility of different problem-constructions has consequences for environmental outcomes as well. When Brazilian presidents speak about the Amazon it not only makes headlines, nationally and internationally (@brice2021; @harris2021; @miranda2021), but also incites responses, shapes expectations, and feeds into the behavior of many actors involved in the Amazon. Deforestation rates in Brazil are more responsive to the government’s environmental policy, generated from material and discursive governmental practices, than exogenous factors as market fluctuations (@assuncao2015; @capobianco2019; @capobianco2021). Ethnographic research has also shown that expectations about governments’ response are a crucial factor in decisions to deforest at the ground (@campbell2015).

@@

Closely related to economic integration, social development is a problem-construction that is often neglected while describing policy cycles is social development. Governments can emphasize the lack of hospitals, sanitation, and schools concerning peoples’ dignity, standards of living, and other constitutional rights when speaking about the Amazon. This is not only relevant as a construction, but speaks, minimally, to democratic concerns over providing electorates, especially in the Amazonian region, with better standards of living than previously.

@@

To analyze how presidents construct the Amazon as a problem in time, we build upon the dataset provided by (Cezar 2020) which contains all official speeches by Brazilian Presidents from 1985 to 2019 scrapped from the archives of the Brazilian Presidential Library. We update the dataset by scraping and adding all official speeches from 2020 and 2021. The final dataset encompasses 6130 speeches for all the presidents of Brazil. Then, we proceed to identify all speeches that refer to the Amazon as a region, people, or forest in these speeches. We do so by detecting all speeches in which the stem “amazon” appears. In Portuguese, the stem captures terms such as “Amazonia”, “Amazonica”, “Amazonidas”, “Amazonense(s)”, “Amazonas”, among others. We find that 946 speeches are, at least partially, about the Amazon from the 6130.

@@

We develop a codebook to code Amazonian statements in one or more problem-constructions (see codebook in appendix). In their conceptualization, each problem-construction is mutually exclusive, meaning that they cover different forms of constructing the Amazon as a problem. Nevertheless, each Amazonian statement might be assigned to one or more codes. A statement can, for example, construct the Amazon as a problem of sovereignty and a problem of economic integration, or a problem of social development and conservation. Amazonian statements, thus, can be either coded as pure-types or mixed-types. Mixed-types are relevant as constructions and policies are often multifaceted while portraying the Amazon as a combination of these problem-constructions. Figure 1, below, portrays this operationalization strategy.

\begin{landscape}

```{r figure1, echo=FALSE, fig.cap="Operationalization of Problem-constructions", out.width = '100%', include=TRUE}
knitr::include_graphics("figure1pic.png")
```

\end{landscape}

@@

We chose to employ a support-vector machine (SVM) algorithm to label texts, that is, a non- probabilistic linear classifier that classifies documents by assigning points in mapped space to maximize the gap between binary categories (Meyer et al. (2021); Noble (2006)) . The SVM model is first trained using the hand coded training set and then employed to classify observations in the hand coded validation set. The trained SVM model was, on average 82%, accurate in classifying observations in the validation set before being tuned. After the SVM model is tuned, we use the model to automatically code the remaining 1007 Amazonian statements. The final dataset for analysis, excluding false positive matches, contains 1895 coded Amazonian statements. Finally, we extract locations for all speeches in the data. These locations represent the Brazilian state in which certain speech was given or an international country.

@@

The frequency in which the Amazon appears across all official presidential speeches varies in time. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches per year. We observe five local maxima: 1989, 1992, 2009, 2015, and 2019. These points, usually, coincide with events that help us explain the rises and falls of the Amazon in presidential discourse in time. For instance, in 1985 the Amazon appeared in about 4% of all the presidential speeches but by 1988 this had increased to around 14%. This is the period when the Brazilian Constitution was written. Indigenous and traditional populations were instrumental in advocating for constitutional environmental rights and the protection of their territories (Hecht and Cockburn 1990). These were eventually enshrined in article 225, which gives all Brazilians a right to a balanced environment, and in article 231, which grants indigenous and traditional populations a right over their territory. However, in 1989, the Amazon appeared in 32% of all speeches. This spike is likely explained by the brutal murder of Chico Mendes and the very end of 1988 (Simons and Times 1988). An incident that caught unprecedented international attention and Sarney responded to this with a set of policies to address deforestation (Capobianco 2021).

```{r Figure 2, fig.cap="Amazonian speeches by year", echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE, include=TRUE}

@@

  geom_text(aes(x = 1988.7, y = 0.34, label = "Chico Mendes"), size = 3) +

@@

  geom_vline(xintercept = c(1988.7, 1992, 2009, 2015),

@@

In 1990 the Amazon appeared in 12% of all speeches and, by 1992, this increased to 25% of all speeches. The 1992 Earth Summit, held in Brazil brought international attention to environmental topics in the Amazon. One of the big announcements, for example, was the consolidation of the first transnational partnership for the Amazon, the G7 Pilot Programme, which brought a high number of financial resources to the region for public policy implementation (Capobianco 2021). Throughout the 1990s and 2000s, during the Cardoso and Lula administrations, the Amazon appeared, on average, on 15% of all speeches. Exceptionally, in 2009 we see another spike in the Amazon appearing in official speeches. This spike coincides with the 2009 Copenhagen Summit as well as the steepest decrease in deforestation rates since data is available. Lula led the delegation to Copenhagen with a self-image of “we do not promise, we deliver” when the stakes about climate change were high (Franchini and Viola 2019).

From 2010 to 2018, with the exception of 2015, we see a general decrease in mentions of the Amazon in all official presidential speeches. From 2010 to 2014, mentions of the Amazon went down from 18% to 8% in official speeches. Disagreements related to the priority of environmental preservation over economic development, most notably in the case of Itaipu Dam, led the environmental minister Marina Silva to resign and run for the presidency on her agenda in 2010. In 2011, the New Forest Code, which regularized land ownership of many illegally deforested areas, was also approved. These are also the years when Brazil entered a long period of political and economic instability, which eventually led to the impeachment of Dilma Roussef. Though, in 2015, the Amazon appeared in 19% of speeches. This spike coincides with the Paris summit which became a key turn in climate politics after the failures of Copenhagen. Brazil went to the Paris Summit with deforestation numbers slightly higher than Copenhagen and a perception that there was a turn towards less conservation. By 2016 mentions of the Amazon in official speeches went down to 6%, the lowest share since 1985.

We subsequently observe a steady increase from 6% in 2016 to almost 24% in the first year of Bolsonaro’s presidency, 2019. At the time, international and national media brought unprecedented attention to Brazilian environmental issues as the record burning of the Amazon, and of the red sky afternoon in São Paulo circulated on social media and international media outlets (cite). This brought issues related to environmental conservation as high priorities in Bolsonaro’s governing agenda with efforts to dismantle key pillars of Brazilian environmental governance (cite). At the same time, President Bolsonaro retrieves Brazil’s hosting status for COP25. In short, relevant international events drive interest and alter how pressing the “Amazon problem” is perceived to be in society, therefore, cause a momentary shift of interest from top policymakers (see Hirschman 1963).

@@

We conceptualize four pure problem constructions: sovereignty, economic integration, social development, and conservation. At the level of the speech, though, presidents might mix multiple problem construction in an Amazonian statement. Pure problem constructions dominate, with an average of  56% of all amazonian statements across time. As above, we observe a strong variation over time for most of these constructions. Figure 3, below, illustrates the share of each pure type in time.
